---
title: "Text_Mining_Assignment"
format: html
editor: visual
---

# Introduction

TODO: rewrite when finished and conclusions

Data: https://data.mendeley.com/datasets/3t9vbwxgr5/3

The dataset, analyzed in the paper *"Temporal Analysis and Visualisation of Music"*, contains music metadata and lyrics from songs released between 1950 and 2019. It includes features like sadness, danceability, loudness, and acousticness, with lyrics. Audio data was collected using the Echo Nest® API via the Spotipy Python package, while lyrics were retrieved through the Lyrics Genius® API based on song titles and artist names.

In this project, we explore how song lyrics have evolved from 1950 to 2020 using a range of text mining techniques. Our goal is to uncover emotional and thematic trends across decades. We analyze song lyrics from 1950-2020 to see if themes have changed.

We use a dataset of English-language song lyrics and apply three main text analysis techniques:

1.  MIAU
2.  **Sentiment analysis**: to track emotions like joy, sadness, anger, and trust over time.
3.  **Bigram analysis**: to see which word pairs are most common in different decades.
4.  **Topic modelling**: to identify the key themes (topics) that dominate lyrics in different eras.

The ultimate goal is to combine these insights to tell a story of musical evolution: TO DO

-   e.g. from expressions of love and romance in the 1950s to the rise of anger, protest, and social commentary in more recent years.

## Libraries

To work with this script, the libraries below are requiered. If not dowloaded yet, set the following chunk to eval=T.

```{r packages downloads, eval=F}
# Install packages if needed (only run these lines once)
install.packages("tidyverse")
install.packages("tidytext")
install.packages("topicmodels")
install.packages("tm")
install.packages("quanteda")
install.packages("widyr")
install.packages("wordcloud")
install.packages("ggraph")
install.packages("igraph") 
```

Now you can load the libraries as normal:

```{r libraries}
library(readr)        # for reading data
library(tidyverse)    # data wrangling and plotting
library(tidytext)     # text mining
library(topicmodels)  # topic modelling (LDA)
library(tm)           # text mining functions
library(quanteda)     
library(widyr)        # pairwise word analysis
library(wordcloud)    # wordclouds
library(ggraph)       # graph visualisation
library(igraph)       # network analysis
library(wordcloud)
library(reshape2)
library(quanteda)
library(quanteda.textplots)
library(forcats)
library(ggplot2)
```

Finally, we load the dataset that we will use.

```{r}
df <- read_csv("tcc_ceds_music.csv")
```

# Initial exploration

Before starting any analysis, to make sure that our dataset is clean and valid for text mining tasks and will work with our functions, we have to:

-   Preview the data structure
-   Check for duplicates
-   Identify any missing values

```{r}
dim(df)
head(df)
names(df)
```

The dataset we are using has over 28372 entries described with 31 variables, such as `artist_name`, `track_name`, `realease_date` or `genre`. We find our most important column under `lyrics`, as this is the column that will give us our text components to analyse.

```{r}
df %>%  
  duplicated() %>%  
  sum()

df %>%  
  summarise_all(~sum(is.na(.)))
```

We find very no duplicates or missing values, which allows us to proceed without major preprocessing.

## Descriptive analysis

Before diving into text processing, we begin with a basic exploration of the dataset to understand its structure and time coverage.

```{r}
# Decades with songs
unique(df$release_date)
length(unique(df$release_date))
2019 - 1950

# Number of songs released per year
df %>% 
  count(release_date) %>% 
  ggplot(aes(x = release_date, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of Songs per Year",
       x = "Year", y = "Number of Songs") +
  theme_minimal()
```

The dataset includes songs from 1950 on onwards until 2019, without skipping any year. The number of songs increases significantly starting in the late 1950s, with a steady rise through the 1980s and into the 21st century. The increase is also visible after 2010, possibly reflecting an improved availability of digital music records.

```{r}
# Top 10 most frequent artists in the dataset
df %>% 
  count(artist_name, sort = TRUE) %>% 
  top_n(10) %>% 
  ggplot(aes(x = reorder(artist_name, n), y = n)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Artists by Number of Songs",
       x = NULL, y = "Number of Songs") +
  theme_minimal()

names(df)
```

```{r}
# Get top 10 artists by number of songs
top_artists <- df %>% 
  count(artist_name, sort = TRUE) %>% 
  top_n(9, n) %>% 
  pull(artist_name)

# Filter dataset to include only those top artists
df_top_artists <- df %>% 
  filter(artist_name %in% top_artists)

# Identify top 10 artists
top_artists <- df %>% 
  count(artist_name, sort = TRUE) %>% 
  top_n(10, n) %>% 
  pull(artist_name)

# Filter dataset
df_top_artists <- df %>% 
  filter(artist_name %in% top_artists)

# Plot: number of songs per year per artist
ggplot(df_top_artists, aes(x = release_date)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ fct_reorder(artist_name, release_date, .fun = min), scales = "free_y") +
  labs(title = "Song Release Distribution of Top 9 Artists",
       x = "Year", y = "Number of Songs") +
  theme_minimal()

# TODO: WITH COLOUR? 
# Filter and summarise data by year
df_top_artists_years <- df %>%
  filter(artist_name %in% top_artists) %>%
  group_by(artist_name, release_date) %>%
  summarise(songs = n(), .groups = "drop")

# Plot: now fill works because we summarised by release year
ggplot(df_top_artists_years, aes(x = release_date, y = songs, fill = release_date)) +
  geom_col() +
  facet_wrap(~ fct_reorder(artist_name, release_date, .fun = min), scales = "free_y") +
  scale_fill_gradient(low = "skyblue", high = "navy", name = "Release Year") +
  labs(
    title = "Song Release Distribution of Top 10 Artists",
    x = "Release Year",
    y = "Number of Songs"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The artists with the most entries are icons like Johnny Cash, Ella Fitzgerald, and Dean Martin. These figures indicate a strong representation of classic country, jazz, and pop artists, likely shaping the lyrical content of earlier decades. There are a few likely reasons why older artists dominate this list. We believe it could be because they have had more time to put songs out. For example, Dolly Parton is still putting music out, yet started in 1967 with her album "Hello, I'm Dolly." The older artists might have also had higher prolific output, as many classic artists had long productive careers with hundres of songs. In comparison, a raper from the 2010s might just be starting to put as many songs out. And since not all become very famous or repected, they might also not be represented in this list. Hence, we also could be dealing with a certain curation bias for established songs. It is very important to appreciate this potential bias, as it may shape the kinds of themes, emotional tones, and language that dominate the lyrics if we were to analyse them without any further distinction.

## Decade Assignment

To account for these generational imbalances and allow for fairer comparisons, we group the songs by decade, enabling us to analyse trends in lyrics over time while controlling for differences in artist representation and output volume.

```{r}
# Create decade variable
df <- df %>% 
  mutate(decade = floor(release_date / 10) * 10) 

# Preview of new column to make sure it works 
df %>% 
  select(release_date, decade) %>%
  sample_n(5)

# Number of songs per decade
df %>% 
  count(decade) %>% 
  ggplot(aes(x = decade, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of Songs per Decade",
       x = "Decade",
       y = "Number of Songs")

df %>% 
  group_by(decade) %>% 
  summarize(total = sum(n()))


```

We observe a steady increase in the number of songs across decades, with the 2010s showing the highest volume (5631 songs). This suggests that the dataset is skewed towards more recent decades. As said before, this likely is due to better online availability of moden music. However, while there are more songs are present of recent decades, these could be more distributed across a wider range of artists. This could explain why the Top 10 Artists chart is dominated by older artists from earlier decades, as they may have had more songs included per person; whereas the 2000s and 2010s may contain a broader and more diverse pool of musicians, each with fewer entries on average.

Having understood the structure of our dataset, we now turn to the lyrics themselves. In the next sections, we apply text mining techniques to uncover patterns in language, emotion, and theme over time.

# Word frequency analysis

To begin our text mining exploration, we break the lyrics into individual words (tokens) and remove common stop words like “the”, “and”, or “of” that carry little semantic weight. We do this by using `anti_join` and a specialized list of stopwords, as stopwords can have high and low frequencies, so filtering out frequencies would not work. This allows us to focus on the most meaningful language used across different decades.

```{r}
lyrics_words <- df %>% 
  unnest_tokens(word, lyrics) %>% 
  anti_join(stop_words) %>% 
  rename(song_id = "...1") # we rename song_id column for clarity

lyrics_words
```

A quick look at the first resulting words, such as "time," "untrue," or "tear," shows they are indeed lyrics-related, emotionally loaded, and thematically rich—suggesting that the tokenisation process has worked as expected.

Now, we turn to what these words can tell us: what were the most common words across decades, and how have songs changed in length over time?

## Words over time

We are looking towards insights that reveal how have the words being used have shifted over time and if they reflect content evolution or genre popularity changes. Even though rudimentary, a simple word count can be a great tool to see how the lyrics, and hence, mood of music has changed over time.

The tables below show the most common words by decades and the most common words overall. The plot below shows the ten most frequently used non-stop words in song lyrics across each decade from the 1950s to the 2010s.

```{r}
most_common_word_by_decade <- lyrics_words %>%
  count(decade, word, sort = TRUE) %>%
  group_by(decade) %>%
  slice_max(n, n = 10)

most_common_word_by_decade

most_common_word_by_decade %>% 
  ungroup() %>% 
  select(-decade) %>%
  group_by(word) %>% 
  summarise(total_n = sum(n)) %>%
  arrange(desc(total_n))

# Plot 
most_common_word_by_decade %>%
  ggplot(aes(reorder_within(word, n, decade), n, fill = decade)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~decade, scales = "free") +
  scale_x_reordered() +
  coord_flip()

```

TODO: Words by decade but colour coded by sentiment??

In the plot we can see that some words appear consistently in the top 10 across all or nearly all decades. "Time" is present in all seven decades, and the most frequent word overall with 26504 appearances. Its consistent presence suggests an enduring focus on temporality concepts that could be memories, moments, urgency, or have a philosophical reflection. In any case, "time" is highly important. The next 7 most frequent words round between the 16700 and 12500 appearences, and include "**feel**," "**heart**," "**yeah**," "**life**," "**baby**," "**leave**," and "**live**." These words suggest that emotion ("feel," "heart"), relationships ("baby," "leave"), and experiences ("live," "life") are timeless themes in music, though the context in which they are used may evolve.

Still, even with common themes, we can observe a noticeable thematic evolution in language:

-   1950s: seem to be dominated by by deeply emotional and romantic terms: "**heart**," "**blue**," "**baby**," "**kiss**," "**feel**," and "**dream**." This marks a lyrical landscape focused on love and idealised romance, typical of pop, blues, soul and country genres. However, they can also be reflective of the growing rock and roll movement, whose most famours representative in this decade was the rock-idol Elvis Presley.

-   1960s: words like "**yeah**" and "**live**" begin to emerge alongside traditional terms, which may reflect a cultural shift towards youth expression, rebellion, and the solidifying rock and roll movement, especially with The Rolling Stones becoming a mayor musical leader in this decade. Of course, The Beatles, their conceptual counterpart, had their golden era also in this decade, which reflects on the persistence on sweet and classical topics of love.

-   1970s: we see a balance between emotional (e.g. "**heart**," "**feel**") and existential ("**life**," "**world**") themes, which could indicate the popularity of disco music famous in the 70s but also the start of more introspective lyrics, perhaps influenced by folk, rock (Aerosmith had seven studio albums chart in the 1970s), and emerging soul/funk genres[^1].

-   1980s: very similar to the 1970s, but with increasing prominence of "**yeah**" and "**baby**," possibly reflecting the rise of indie pop, synth-pop and metal. At the end of the decade hip hop would bloom in its gold age which would last until the early 90s. However, the topics did not change much, which might indiciate very universal topic selection by the genres artists

-   1990s: the list becomes more balanced and stable, with no strong newcomers but a continued emphasis on emotion, energy ("**yeah**," "**live**") and relationships, which could be attributed to the continuation of popularity of various types of pop from the decade before. Notably, electronic music, house and techno grew highly popular in this decade, which could explain the lack of introduction of new genre-specific words, as the music has a liminal reliance on lyrics.

-   2000s: here we see the introduction of new casual expressions: "**gonna**" and "**break**" enter the top 10, indicating a more conversational, assertive tone and a bolder lyrical style. This possibly due to the influence of teen pop, with renownded artists such as \*NSYNC, Backstreet Boys or Britney Spears as representatives. R&B was also one of the most popular genres, showcased by the emergence of Beyoncé, Rihanna and Alicia Keys[^2].

-   2010s: we witness a wew standout word with "**fuck**," which is the only profanity in the top list, marking a significant shift in lyrical boldnes (but not necessarily social acceptability, as these type of lyrics are still highly debated). It also includes "**wanna**," reinforcing the trend toward casual and direct language. These changes point to success of adult contemporary music (Bruno Mars, Maroon 5, Taylor Swift), the continuous growth of hip hop and rap (Kendrick Lamar, Future, Chanxe the Rapper, Nicki Minaj), and the rise of trap and explicit pop (Young Thug, 21 Savage, Migos) where raw, unfiltered expression is more common.

[^1]: [Wikipedia](https://en.wikipedia.org/wiki/1970s_in_music) has great articles on the popular music throughout the decades, such as for the 1970s: "In North America, Europe, and Oceania, the decade saw the rise of disco, which then went on to become one of the biggest genres of the decade, especially in the mid-to-late 1970s. In Europe, a variant known as Euro disco rose in popularity towards the end of the 1970s. Aside from disco, funk, soul, R&B, smooth jazz, and jazz fusion remained popular throughout the decade. Rock music played an important part in the Western musical scene, with punk rock thriving throughout the mid to late 1970s."

[^2]: [Wikipedia](https://en.wikipedia.org/wiki/2000s_in_music): "In 2004, the Billboard Year-End Hot 100 had 15 of its top 25 singles as contemporary R&B."

All in all we can say that these patterns reflect both musical genre shifts and changing social and cultural norms. The lyrical focus of songs has expanded and modernised over time. As the early decades (1950s–1970s) are steeped in romantic and emotional vocabulary, we see that starting in the 1980s and especially in the 2000s and 2010s, we see more casual, assertive, and even explicit language emerging, reflecting broader cultural changes around self-expression, social norms, and genre evolution. Finally, the consistency of emotional words like "heart," "feel," and "baby" shows that, despite these shifts, music has always remained a vehicle for emotional storytelling.

Having seen now the emotional side of word frequency analysis, we explore now a more technical question: are songs getting longer or shorter?

## Structural decomposition of songs

In the table we can have the evolution of song length since the 1950s, which we have visualised in the graph.

```{r}
lyrics_stats <- lyrics_words %>%
  group_by(song_id, decade) %>%
  summarise(total_words = n()) %>%
  group_by(decade) %>%
  summarise(mean_length = mean(total_words))

lyrics_stats

# Plot
ggplot(lyrics_stats, aes(x = decade, y = mean_length)) +
  geom_point(aes(colour = as.factor(decade)), size = 8) + 
  scale_colour_viridis_d(name = "Decade", option = "C") +
  scale_y_continuous(limits = c(0, NA)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average lyric length per song by decade (1950–2010)",
    x = "Decade",
    y = "Average Word Count"
    ) + 
  theme_minimal()
```

We can see that lyric length remained relatively stable in the 1950s and 1960s, but then began a steady rise from the 1970s onward. By the 2010s, the average song had over 70 words, roughly 60% more than in the 1950s.

This upwards trend may reflect a move toward more lyrically dense genres like rap and indie pop, changes in song structure, with more verses and fewer instrumental breaks, or maybe a change in production norms allowing for longer tracks.

However, while lyrics have become longer over time, this does not necessarily mean they’ve become more complex: using more words does not always mean saying more. It could simply reflect a shift in lyrical style or vocabulary.

While word frequency tells us what kinds of words dominate lyrics, it doesn't reveal the emotional tone behind them. To better understand what is being expressed, rather than just how much, we now turn to sentiment analysis to explore the emotional tone of songs across decades.

# Sentiment Analysis

Sentiment Analysis is a great tool to quantify the emotions being expressed in our lyrics.

## TODO: what the hell is this graph ??

```{r}
sentiment_by_decade <- lyrics_words %>% 
  inner_join(get_sentiments("bing")) %>% 
  group_by(decade, sentiment) %>%
  count() %>% 
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment_score = positive - negative)

ggplot(sentiment_by_decade, aes(x = decade, y = sentiment_score)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Net Sentiment per Decade",
       y = "Sentiment Score (pos - neg)")

```

## Emotions in lyrics across time

Firstly, we use the NRC lexicon, which classifies words into categories like **joy**, **sadness**, **anger**, **trust**, and **fear**. By mapping each word in the lyrics to these emotional labels, we can track how the emotional landscape of music has changed over time.

```{r}
# sentiment attribution 
nrc_emotions <- lyrics_words %>% 
  inner_join(get_sentiments("nrc")) %>% 
  count(decade, sentiment) %>% 
  filter(sentiment %in% c("anger", "joy", "fear", "trust", "sadness"))
nrc_emotions

# Plot
ggplot(nrc_emotions, aes(x = decade, y = n, fill = sentiment)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
  "anger" = "#e21818",
  "fear" = "#976cb6",
  "sadness" = "#54bbdd",
  "joy" = "#ffee35",
  "trust" = "#76dd54"
)) +
  labs(title = "Emotions per Decade (NRC)",
       y = "Count",
       x = "Decade") + 
  theme_minimal() 

```

This plot shows the count of emotion-related words found in lyrics across each decade (1950–2010) using the NRC sentiment lexicon. As expected, the total number of emotion-related words increases steadily over the decades, which reflects the growing number of songs in the dataset and the longer lyrics by song in recent decades. With this plot we can analyse the relative prominence and pace of change between emotions.

**Sadness**, **fear**, and **trust** are the most prominent emotional categories and show particularly sharp growth from the 1980s onwards. Their rise suggests that music has become increasingly reflective, vulnerable, and emotionally rich over time. **Anger**, although less frequent than the other emotions in earlier decades, sees a noticeable surge in the 2000s and especially the 2010s, more than doubling in count. This may reflect a cultural climate marked by frustration, protest, or disillusionment.

Interestingly, while the count of words associated with **joy** also increases, its growth is more modest in comparison. By the 2010s, joy appears to be outpaced by sadness, fear, and even trust. This could indicate a relative decline in cheerful or celebratory themes in mainstream music, possibly due to the rise of genres like hip hop, alternative rock, and indie pop that tend to focus more on personal struggle, identity, and existential topics.

However, if we actually want to analyse the emotions by decade independently of the selection bias of our dataset, we have to take a different approach. A more robust tactic is to calculate the mean number of emotion-related words per song within each decade. With this we help control for selection bias and can more reliably make cross-decade comparisons. The goal is to shift the focus from overall volume to lyrical density: how emotionally expressive the average song is in each decade, independent of how many songs were written.

```{r}
# sentiment attribution
nrc_emotions <- lyrics_words %>%
  inner_join(get_sentiments("nrc")) %>%
  count(decade, sentiment)

# number of songs per decade
songs_per_decade <- df %>%
  group_by(decade) %>%
  summarise(song_count = n())

# join + mean per song 
nrc_emotions_normalised <- nrc_emotions %>%
  left_join(songs_per_decade, by = "decade") %>%
  mutate(mean_per_song = n / song_count) %>%
  filter(sentiment %in% c("anger", "joy", "fear", "trust", "sadness"))
nrc_emotions_normalised

# Plot
ggplot(nrc_emotions_normalised, aes(x = decade, y = mean_per_song, fill = sentiment)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "anger" = "#e21818",
    "fear" = "#976cb6",
    "sadness" = "#54bbdd",
    "joy" = "#ffee35",
    "trust" = "#76dd54"
  )) +
  labs(
    title = "Mean Emotion per Song by Decade (NRC)",
    y = "Mean Emotion per Song",
    x = "Decade",
    fill = "Emotion"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

By normalising the number of emotion-related words per song, we avoid bias from having more songs in recent decades. The data reveal a more moderate shift in emotional expression over time.

In the 1950s and 60s, lyrics were most commonly associated with **joy** and **trust**, while **anger** and **fear** appeared far less frequently. However, from the 1980s onwards, we observe a steady rise in words related to **sadness**, **fear**, and **anger**. By the 2010s, **sadness** and **fear** have become the most dominant emotional categories per song, with anger also reaching its highest levels since it more than doubled between the 1950s and 2000s.

Although **joy** remains relatively stable across decades, it follows a very different patterns than the other emotions, as it decreases from the 1950s to the 1960s, and then mantains a steady course onward. This implies that while positive sentiment continues to exist in the lyrics, it is increasingly balanced or even overshadowed by more complex and emotionally intense language.

Going back to the methodology, compared to the earlier analysis based on total emotion word counts, this normalised view reveals a more nuanced picture. While the previous plot suggested a steady, almost uniform growth in all emotions, the per-song averages show that emotions like sadness and fear have grown more sharply than others. This could mean that the emotional intensity of the average song has increased, especially for darker or more negative sentiments. Modern lyrics, then, appear not just longer but also more diverse, complex and richer in emotional depth.

Due to the prominent rising of the anger emotion, we take a look at its expression over the decades.

```{r}
# sentiment selection 
nrc_anger <- get_sentiments("nrc") %>% 
  filter(sentiment == "anger")

# sentiment extraction by decade
anger_words_by_decade <- lyrics_words %>%
  inner_join(nrc_anger, by = "word") %>%
  count(decade, word, sort = TRUE)

# grouping by decade for table 
top_anger_words_by_decade <- anger_words_by_decade %>%
  group_by(decade) %>%
  slice_max(order_by = n, n = 10) %>%
  arrange(decade, desc(n)) %>%
  ungroup()

top_anger_words_by_decade

# plot 
anger_words_by_decade %>% 
  group_by(decade) %>% 
  top_n(10, n) %>%  
  ungroup() %>% 
  ggplot(aes(x = reorder_within(word, n, decade), y = n, fill = factor(decade))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ decade, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "Words associated with 'anger' by decade",
     x = "Word",
     y = "Frequency") + 
  theme_minimal()
```

The most common anger-related words in earlier decades, like "**lose**," "**lonely**," and "**hurt**," reflect a more emotional and personal tone. By the 1970s and 80s, words such as "**fight**," "**shoot**," and "**hell**" become more frequent, signalling a shift toward more confrontational expressions of anger. In the 1990s and especially the 2010s, explicit terms like "**shit**," "**bitch**," and "**damn**" rise sharply, reflecting the influence of hip-hop and a broader acceptance of raw, unfiltered language in music. This evolution points to a shift from internalised frustration to more direct, aggressive expressions of external anger in modern lyrics.

Let us see how these anger, as well as the other emotions, change in frequency over decades.

```{r}
# FIND EMOTIONS ---------------------------
# joy -------------------------------------
# sentiment selection 
nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy")

# sentiment extraction by decade
joy_words_by_decade <- lyrics_words %>%
  inner_join(nrc_joy, by = "word") %>%
  count(decade, word, sort = TRUE)

# sentiment extraction (independent of decade)
top_joy_words <- lyrics_words %>%
  inner_join(nrc_joy, by = "word") %>%
  count(word, sort = TRUE) %>% 
  slice_max(order_by = n, n = 10)

# anger -----------------------------------
# previous steps already done in chunk before 
top_anger_words <- lyrics_words %>%
  inner_join(nrc_anger, by = "word") %>%
  count(word, sort = TRUE) %>% 
  slice_max(order_by = n, n = 10)

# sadness ----------------------------------
nrc_sadness <- get_sentiments("nrc") %>% 
  filter(sentiment == "sadness")

sadness_words_by_decade <- lyrics_words %>%
  inner_join(nrc_sadness, by = "word") %>%
  count(decade, word, sort = TRUE)

top_sadness_words <- lyrics_words %>%
  inner_join(nrc_sadness, by = "word") %>%
  count(word, sort = TRUE) %>% 
  slice_max(order_by = n, n = 10)

# fear --------------------------------------
nrc_fear <- get_sentiments("nrc") %>% 
  filter(sentiment == "fear")

fear_words_by_decade <- lyrics_words %>%
  inner_join(nrc_fear, by = "word") %>%
  count(decade, word, sort = TRUE)

top_fear_words <- lyrics_words %>%
  inner_join(nrc_fear, by = "word") %>%
  count(word, sort = TRUE) %>% 
  slice_max(order_by = n, n = 10)

# trust -------------------------------------
nrc_trust <- get_sentiments("nrc") %>% 
  filter(sentiment == "trust")

trust_words_by_decade <- lyrics_words %>%
  inner_join(nrc_trust, by = "word") %>%
  count(decade, word, sort = TRUE)

top_trust_words <- lyrics_words %>%
  inner_join(nrc_trust, by = "word") %>%
  count(word, sort = TRUE) %>% 
  slice_max(order_by = n, n = 10)

# PLOT ---------------------------------------
# bind top words for all five emotions
top_emotion_words <- bind_rows(
  top_joy_words %>% mutate(sentiment = "joy"),
  top_sadness_words %>% mutate(sentiment = "sadness"),
  top_anger_words %>% mutate(sentiment = "anger"),
  top_fear_words %>% mutate(sentiment = "fear"),
  top_trust_words %>% mutate(sentiment = "trust")
)

# filter for top emotions  
top_emotion_counts <- lyrics_words %>%
  semi_join(top_emotion_words, by = "word") %>%
  count(decade, word) %>% 
  # add sentiment labels 
  left_join(
    top_emotion_words %>% select(word, sentiment), 
    by = "word")

# top_emotion_counts # table

# heatmap
ggplot(top_emotion_counts, aes(x = factor(decade), y = reorder(word, n), fill = n)) +
  geom_tile(colour = "white") +
  facet_wrap(~ sentiment, scales = "free_y") +
  scale_fill_gradient(low = "white", high = "#084594") +
  labs(
    title = "Top Emotion-Related Words by Decade",
    x = "Decade",
    y = "Word",
    fill = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())

```

Taking a look at the heatmap of the most frequent emotionally charged words over time reveals both continuity and change in lyrical expression.

Words like "**baby**," "**sweet**," and "**sing**" (associated with **joy** and **trust**) appear consistently in every decade, pointing to a lasting emotional core in popular music centred around love, tenderness, and musicality.

A couple words like "**lonely**" and "**blue**" under **sadness** show a decreasing trend, in favour of words like "**leave**" and "**lose**," highlighting the enduring nature of heartbreak and isolation as lyrical themes.

On the other hand, terms like "**bitch**," "**shit**," and "**kill**," linked to **anger**, **sadness**, and **fear**, have surged dramatically in recent decades, especially from the 1990s onward. Confirming our previous word count analysis, their rise reflects a shift toward more explicit, raw, and confrontational language.

## Positive and negative sentiment over time

After examining individual emotions and specific words, we now turn to a broader comparison of general sentiment: how the balance between positive and negative language in lyrics has changed over time.

```{r}
# ratio calculation
bing_ratio <- lyrics_words %>%
  inner_join(get_sentiments("bing")) %>%
  count(decade, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 1) %>%
  mutate(ratio = positive / negative)

# table
bing_ratio

# Plot
ggplot(bing_ratio, aes(x = decade, y = ratio)) +
  geom_point(aes(colour = as.factor(decade)), size = 8) + 
  scale_colour_viridis_d(name = "Decade", option = "C") +
  scale_y_continuous(limits = c(0, 0.80)) +
   geom_text(aes(label = round(ratio, 2)), vjust = -1, size = 3.5, colour = "black") +
  geom_line() +
  geom_point() +
  labs(title = "Positive-to-Negative Ratio in song lyrics per Decade",
       x = "Decade", 
       y = "Positive / Negative Word Ratio") + 
  theme_minimal()
```

The graph shows a clear and continuous decline in the ratio of positive to negative words from the 1950s to the 2010s. In the 1950s, there were approximately **66 positive** words for every **100 negative words**, but by the 2010s, this had dropped to just **35 positive** words per **100 negatives**. This shift suggests that lyrics have become increasingly dominated by **negative** sentiment over the past 70 years, reinforcing our previous findings about the growing prevalence of sadness, fear, and anger.

Modern songs not only use more emotionally charged language, but they also skew more negative in tone. The consistent drop across each decade suggests this is not due to fluctuations in genre or artist popularity but it actually reflects a deeper, long-term shift in how artists express themselves. Lyrics today are not only longer and richer in vocabulary and emotion, but also more likely to focus on negative emotions than they were in previous generations.

Lets visualise this trend with some wordclouds.

```{r}
#| warning: false
library(wordcloud)
library(reshape2)

#general
set.seed(123)

# with tidy ------------------------------------------
lyrics_words %>%
  inner_join(get_sentiments("bing"), by = "word") %>%
  count(word, sentiment, sort = TRUE) %>%
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  comparison.cloud(
    colors = c("#D83842", "#88C564"),
    max.words = 100, 
    scale = c(2, 0.3), 
    random.order = FALSE
    )

```

```{r}
library(quanteda)
library(quanteda.textplots)

# TODO 
# with DFM --------------------------------------------
# extract sentiments
sentiment_words <- lyrics_words %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE)

# DFM with counts only once per word, not per decade
dfm_sent <- sentiment_words %>%
  mutate(doc = "all_lyrics") %>%
  cast_dfm(doc, word, n)

# match colours to sentiment
word_sentiments <- sentiment_words %>%
  distinct(word, sentiment)

word_colors <- ifelse(word_sentiments$sentiment == "positive", "#88C564", "#D83842")

textplot_wordcloud(dfm_sent,
                   min_count = 5,
                   random_order = FALSE,
                   rotation = .25, 
                   color = word_colors)
```

The wordcloud visualises the most frequent positive and negative words found in the lyrics, based on the Bing sentiment lexicon.

We see that negative words such as "**break**," "**fall**," "**lose**," "**shit**," "**fuck**," and "**bitch**" appear prominently for the negative words, reflecting the strong presence of emotional conflict, struggle, and explicit language in modern music. Positive terms are exemplified by "**sweet**," "**love**," "**free**," and "**ready**". However, only "**love**" and "**sweet**" are of considerable size, while the other positive words seem to be generally rather smaller, signifying their less frequent appearances. This echoes what we observed in the positive-to-negative ratio plot.

## Conclusion

Overall, sentiment analysis reveals a clear emotional evolution in lyrics across the decades, with a growing presence of negative sentiment and emotionally intense language in more recent songs, suggesting that contemporary music is increasingly used as a space for expressing inner struggles, social tension, and complex emotions.

Having explored the emotional tone of lyrics thorugh sentiment analysis, we take our analysis now one step further to look at patterns in word pairings.

# N-GRAMS

N-grams allow us to examine not just individual word frequencies, but also how words are used together. To goal is to uncover recurring phrases, expressions, and linguistic patterns in lyrics, revealing themes, styles, and relationships between concepts.

We choose to analyse bi-grams, as they can reflect idiomatic phrases, emotional expressions, or genre-specific language patterns. With count we can examine the most frequent bi-grams.

```{r}
songs_bigrams <- df %>% 
  unnest_tokens(bigram, lyrics, token = "ngrams", n = 2) %>%
  filter(!is.na(bigram))

songs_bigrams %>% 
  select(bigram, decade) %>%
  sample_n(5)

songs_bigrams %>% 
  count(bigram, sort = TRUE)
```

Because many of the most frequent bigrams consist of common stopwords, we proceed to clean the bigram data by removing stopwords from both word positions and also create a dataset that counts the bigrams by decade.

```{r}
bigrams_separated <- songs_bigrams %>%
  separate(bigram, c("word1", "word2"), sep = " ")

bigrams_filtered <- bigrams_separated %>% 
  filter(!word1 %in% stop_words$word) %>% 
  filter(!word2 %in% stop_words$word)

bigrams_united <- bigrams_filtered %>% 
  unite(bigram, word1, word2, sep = " ")

bigram_counts_decade <- bigrams_united %>% 
  count(decade, bigram, sort = TRUE)

```

Having prepared our various bigrams datasets, we can now do some analysis.

## Follow up words to Love and Hate

W want to explore conditional n-grams, i.e. how the second word in a phrase depends on the first to better understand how certain emotionally charged words are used in context.

Let's see what is the most repeated word after love in these songs through the decades:

```{r}
love_bigrams <- bigrams_filtered %>%
  filter(word1 == "love") %>%
  count(decade, word1, word2, sort = TRUE) 

# love_bigrams

bigrams_filtered %>%
  filter(word1 == "love") %>%
  count(decade, word2, sort = TRUE) %>%
  group_by(decade) %>%
  slice_max(order_by = n, n = 5) %>%
  ggplot(aes(x = reorder_within(word2, n, decade), y = n, fill = factor(decade))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~decade, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(
    title = "Top words following 'love' by decade",
    x = "Following word",
    y = "Frequency"
  )

```

In all decades, the most popular word after "**love**" is "**love**." This might be explained by pop music and chorus with repetitive lyrics, e.g. "love, love, love."

Using quanteda, we can take a quick look at the context of the word love.[^3]

[^3]: If time is available, we highly recomend looking at lyrics with the word "crab" in it. They are incredible. Do not ponder how we stumbled across it.

```{r}
library(quanteda)

lyrics_corpus <- corpus(df$lyrics, docvars = data.frame(label = df$decade))

kwic(tokens(lyrics_corpus), "love", window = 5)

df %>%
  slice(666) 

```

For example, on page 11 for the line of \[text666, 41\] we see that the 10 words in which "**love**" sits between are "**heart want love love love**" and "**look eye time realize love**." Taking a look at the specific song we see that it is a 1963 song called "I'm the one who loves you" by "The Impressions," and the specific sentences goes like this:

"Also confess what my **heart** **wants** me to\
That I'm the one who **loves** you\
I'm the one who **loves** you\
A **love** me\
I'm the one who **loves** you\
Yes, once more you'll **look** in my **eyes**\
And one more **time** you won't **realize**\
That I'm the one who **loves** you"[^4]

[^4]: <https://genius.com/The-impressions-im-the-one-who-loves-you-lyrics>

This is a great example to see how the word "**love**" becomes so overly when conditioned by the previous word "**love**." Through the work of filtering stopwords we see how actually there is quite a lot of context between each love, but yet the stopword filtering reveals the central and undisputable topic of the lyrics: lots of love.

Other interesting words after "**love**" are "**eternally**" (1950s), "**girl**" (1960s), "**mambo**" (1970s), "**fool**" (1980s), "**dead**" and "**woooy**" (1990s), "**savior**" (2000s), "**arm**" and "**crew**" (2010s).

Let's look now at words after "**hate**."

```{r}
bigrams_filtered %>%
  filter(word1 == "hate") %>%
  count(decade, word2, sort = TRUE) %>%
  group_by(decade) %>%
  slice_max(order_by = n, n = 5) %>%
  ggplot(aes(x = reorder_within(word2, n, decade), y = n, fill = factor(decade))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~decade, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(
    title = "Top words following 'hate' by decade",
    x = "Following word",
    y = "Frequency"
  )
```

The conditional analysis for the word hate strongly reiterates our previous findings. In the 1950s and 60s, "**hate**" is mostly followed by personal or emotional words like "**lose**," "**guess**," or "**hear**", suggesting internal or relational struggles. Notably, "**school**" appears in the 1960s. By the 1970s and 80s, themes broaden slightly, with words like "**bomb**" and "**music**", hinting at more societal or cultural references.

The 1990s see a extremely sharp rise in the repetition of "**hate** **hate**", indicating a more intense tone. In the 2000s and 2010s, the double repetition of "**hate**" continues, accompanied by expressions like "**tear**," "**machine**," "**world**," "**leave**," and "**stop**" show growing existential and emotional complexity.

## Double negation

A key limitation we encountered during our analysis relates to the treatment of negation. For example, while the word "love" is generally classified as positive, its meaning can be retracted when preceded by "not." Hence, the analysis of bigrams with negation words is useful for this problematic.

However, as our analysis progressed we began to suspect that our dataset may have been pre-cleaned. This culminated in this section, where we specifically are unable to find any trace of common negation words such as "no," "never," "not," or "without" in any of the lyrics. This absence strongly suggests that the dataset’s publishers removed stopwords (including negations) prior to publication, which limits our ability to examine how these negations affect sentiment and meaning in context. Sadly, we were not aware of this until this point, as the published did not disclose this information in their dataset description.

```{r}
# :( 
df %>%
  filter(str_detect(
    str_to_lower(lyrics),
    "\\bnever\\b|\\bnot\\b|\\bno\\b|\\bwithout\\b"
  ))
```

## Moods and themes

Beyond general bigram frequencies, we now narrow our focus to a set of curated phrases that reflect dominant themes in music lyrics: love and relationships, partying and escapism, violence, and social awareness. These specific bigrams were chosen due to their commonality in popular music and because they are emotionally and culturally charged. By tracking their use over time, we can observe how the focus of lyrical content has shifted across decades.

To capture these bigrams in their original form, we use the version of the dataset that retains stopwords. As we have explained, we actually do not have a list with the full lyrics', but as we do not know exactly what dictionary has been used to remove words, we will still attempt to make the best out of it.

```{r}
bigram_counts_decade <- songs_bigrams %>% 
  count(decade, bigram, sort = TRUE)

selected_bigrams <- c(
  "broken heart", "kiss me", "feel good", #love themes
  "get high", "all night", "dance all", #party themes
  "kill you", "die young", "pull trigger", #violence
  "black lives", "stand up", "wrong system") #social conscience 

# Plot
bigram_counts_decade %>%
  filter(bigram %in% selected_bigrams) %>%
  ggplot(aes(x = factor(decade), y = n, fill = bigram)) +
  geom_col(position = "dodge") +
  labs(
    title = "Selected Bigram Frequency by Decade",
    x = "Decade",
    y = "Frequency",
    fill = "Bigram"
  ) +
  theme_minimal(base_size = 13)
```

The phrase "**feel** **good**" dominates the chart, showing a steady rise from the 1960s onward and with high peack in the 1970s and in the 2010s. Its consistent presence suggests that themes of positivity have remained central to popular music. In contrast, "**get** **high**" and "**pull trigger**", which carry darker and edgier connotations, were virtually absent in the 1950s and 60s, but became noticeably more frequent starting in the 1980s and then later in the 2000s, reflecting the growing integration of drug and violence-related themes into mainstream lyrics. "**Die** **young**" appears much less frequently overall, with scattered usage and no clear upward trend, though its small peak in the 2000s may signal the occasional rise of existential or tragic motifs in pop culture narratives.

Overall, the tall bars for "**feel good**" contrast sharply with the more recent emergence of phrases like "**get high**" and "**pull trigger**", highlighting how lyrical content has shifted from simple feel-good messages to increasingly complex and darker themes overall, which confirms our findings in the sentiment analysis section.

### TF-IDF

After examining our hand-picked bigrams, we now shift to a more systematic approach. TF-IDF helps us identify the most distinctive bigrams in each decade based on their frequency and uniqueness. The goal is to uncover uniquely characteristics of each decade’s lyrical style, trends and themes.

```{r}
bigram_tf_idf <- bigrams_united %>%
  count(decade, bigram) %>%
  bind_tf_idf(bigram, decade, n) %>%
  arrange(desc(tf_idf))

bigram_tf_idf %>%
  group_by(decade) %>%
  slice_max(tf_idf, n = 5) %>%
  ungroup() %>%
  ggplot(aes(tf_idf, fct_reorder(bigram, tf_idf), fill = decade)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~decade, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```

The top-scoring bigrams by decade reveal a wide variety of decade-specific expressions, many of which are stylistically unique, repetitive, or culturally embedded. In the 1950s and 60s, we observe a prevalence of nonsensical, playful, or rhythmic phrases like "**zinga** **zinga**", "**putty** **putty**", and "**gong** **gong**".

By the 1970s and 80s, the most distinctive bigrams begin to take a darker turn. Phrases like "**women** **bleed**", "**church** **poison**", and "**mellow** **dead**" suggest an increasing willingness to explore themes of violence, institutional distrust, and existential reflection. This shift likely mirrors both changing musical genres and broader societal transformations during these decades.

From the 1990s to the 2010s, the vocabulary becomes more politically and emotionally charged. Bigram standouts such as "**real** **terrorist**", "**bleed** **control**", "**wicked** **wicked**", and "**preach** **preach**" point to lyrics that engage directly with identity, power, and social critique. Some phrases (like "**bout** **bass**") also reflect the growing influence of hip-hop and electronic genres, where rhythm, slang, and cultural reference play central roles.

The diversity of top bigrams across decades highlights how lyrics evolve not just in vocabulary, but in tone, themes, and cultural references

## Correlating

Building on our exploration of bi-grams, we wanted to further investigate how words are used together in lyrics—not only as adjacent phrases, but also in terms of broader associative patterns across entire songs. While bi-grams help reveal repeated expressions and fixed structures, co-occurrence and correlation analyses offer a deeper understanding of the conceptual relationships between words. These approaches help us see not just what is said, but how ideas are built and connected throughout a song.

To begin, we used the pairwise_count() function from the widyr package to compute how often two words appear together within the same song. This allows us to move beyond adjacent word pairs and capture thematic links that might not be sequential but still co-exist within the lyrical narrative:

```{r}
library(widyr)

word_pairs <- lyrics_words  |>  
  pairwise_count(word, song_id, sort = TRUE)
```

However, since our analysis is focused on change over time, we wanted to examine how these word relationships vary by decade. We grouped the lyrics by decade and applied the same co-occurrence method to each group:

```{r}
word_pairs_by_decade <- lyrics_words |> 
  group_by(decade) |>  
  group_split() |>  
  map_df(~pairwise_count(.x, word, song_id, sort = TRUE) |> 
           mutate(decade = unique(.x$decade)))

word_pairs_by_decade
```

This gave us a decade-by-decade view of which word pairs were most commonly used together. We filtered the results to include only pairs that appeared at least ten times and selected the top 20 for each decade. These were visualised in a faceted bar chart, highlighting the most prominent co-occurring word combinations and their evolution:

```{r, fig.height= 13}

top_bigrams <- word_pairs_by_decade |> 
  filter(n >= 10) |> 
  group_by(decade) |> 
  slice_max(order_by = n, n = 20) |> 
  ungroup() |> 
  mutate(bigram = paste(item1, item2))  


top_bigrams <- top_bigrams |> 
  group_by(decade) |> 
  mutate(bigram = reorder_within(bigram, n, decade)) |> 
  ungroup()


ggplot(top_bigrams, aes(bigram, n, fill = decade)) +
  geom_col(show.legend = FALSE) +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~decade, scales = "free_y") +
  theme_minimal(base_size = 20) +
  labs(
    title = "Top 20 Word Pairs by Decade",
    x = "Bigram (word pair)",
    y = "Co-occurrence count"
  )
```

This visualization reveals several interesting patterns. In the 1950s and 1960s, bigrams revolving around the word “**heart**” — such as “**heart break**”, “**break heart”**, and “**kiss heart**” — dominate, reflecting the lyrical focus on romantic and emotional themes. As we move into the 1970s and 1980s, we see a shift toward introspection and existential themes, with frequent bigrams like “**time feel**”, “**feel time**” and “**live life**”. These expressions suggest a broader reflection on life, emotions, and personal experiences.

In the 2010s, the nature of lyrical expression seems to evolve significantly. The most common bigrams include phrases like “**lyric commercial**” and “**wanna time**”, indicating not only the rise of meta-lyrical content — lyrics referencing lyrics themselves — but also more colloquial, conversational phrasing. This change could be attributed to the influence of commercial pop culture, hip-hop, and the increasing informality in language use.

While co-occurrence reveals which word pairs are frequent, it does not necessarily indicate a meaningful or strong connection between them. To go one step further, we computed pairwise correlations using pairwise_cor(), which measures how often two words appear in the same songs relative to their individual frequencies. We calculated word correlations based on shared song IDs, filtering for words that appeared at least 20 times to ensure reliability:

```{r}
word_cors <- lyrics_words %>%
  group_by(word) %>%
  filter(n() >= 20) %>%
  pairwise_cor(word, song_id, sort = TRUE)

word_cors
```

We then visualized strong correlations (greater than 0.4) using a correlation network:

```{r}
set.seed(2016)

word_cors |> 
  filter(correlation > 0.4) |> 
  graph_from_data_frame() |> 
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation), show.legend = FALSE) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void()
```

This network plot reveals clusters of words that are semantically or thematically related. For instance, we observe nodes of emotional expressions, genre-specific terms, and even playful or non-standard phonetic sounds like "dingdongding" or "shooby," likely linked to jazz, soul, or novelty songs. The structure of this network highlights how certain words consistently co-occur across various songs, pointing to recurring motifs or lexical choices that characterize particular styles or genres.

Then, to maintain our focus on temporal variation, we repeated the correlation analysis by decade. We grouped the lyrics accordingly and computed the correlations for each group separately:

```{r}
word_cors_by_decade <- lyrics_words |> 
  group_by(decade) |> 
  group_split() |> 
  map_df(~ .x |> 
           group_by(word) |> 
           filter(n() >= 20) |> 
           pairwise_cor(word, song_id, sort = TRUE) |> 
           mutate(decade = unique(.x$decade))
  )
```

This allowed us to create separate word networks for different decades, helping us trace how word associations and thematic clusters have evolved. For example, the following graph illustrates the strongest correlations in the lyrics of the 2010s:

```{r}
word_cors_by_decade |> 
  filter(decade == 2010, correlation > 0.4) |> 
  graph_from_data_frame() |> 
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation), show.legend = FALSE) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void() +
  ggtitle("Word Correlation Network – 2010s")

```

By contrast, this is the same graph for the 1950s:

```{r}
word_cors_by_decade |> 
  filter(decade == 1950, correlation > 0.4) |> 
  graph_from_data_frame() |> 
  ggraph(layout = "fr") +
  geom_edge_link(aes(edge_alpha = correlation), show.legend = FALSE) +
  geom_node_point(color = "lightblue", size = 5) +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void() +
  ggtitle("Word Correlation Network – 1950s")

```

On one hand, the network for the 2010s reveals several distinct and tightly knit clusters, reflecting a noticeable shift toward more explicit and conversational language. Words such as “**fuck**”, “**shit**”, “**bitch**” and “**nigga**” form a densely connected group, indicating the strong presence of profane and direct expressions. Additionally, terms like “**remix**”, “**feat**” and “weeknd” highlight a more self-aware and commercial dimension of music in this decade, where artists frequently reference the music industry, collaboration, and celebrity culture within their lyrics.

Alongside these clusters of explicit language and cultural references, there are smaller groupings of non-lexical vocalizations such as “**lalalala**” or “**mymy**” pointing to a performative use of sound and rhythm beyond semantic meaning.

On the other hand, the 1950s network shows a more cohesive and thematically focused structure. Prominent clusters include words related to sentimentality and romance, such as “**tenderness**”, “**simple**” and “**sentimental**”, which aligns with the emotional tone prevalent in mid-century ballads. Playful, sound-based terms like “**ding**”, “**dong**” and “**whistle**” also appear, reflecting an interest in phonetic play and melodic embellishments typical of early rock 'n’ roll. Interestingly, words like “**fuck**”, “**bitch**” or "**nigga**" do appear in the 1950s network, possibly hinting at the presence of fringe or subversive lyrical content even during a more culturally conservative era.

# Topic modeling

To uncover broader thematic patterns in song lyrics, we turned to topic modelling, a powerful text mining technique that enables us to detect underlying semantic structures in large collections of documents. While earlier analyses focused on surface-level lexical features — such as word frequency, bigrams, or word associations — topic modelling provides a more abstract, yet insightful, lens through which to view the evolution of meaning and content over time.

Specifically, we employed Latent Dirichlet Allocation (LDA), a probabilistic model that assumes each document (in our case, a song or group of songs) is a mixture of multiple topics, and that each topic is itself a distribution over words.

Our first step was to prepare the data. We began by tokenizing the lyrics and removing stopwords to focus only on the meaningful terms. Then, we counted the occurrences of each word per decade, giving us a term frequency distribution that we used to build a Document-Term Matrix (DTM), where each decade represented a document:

```{r}
word_counts <- df %>% 
  unnest_tokens(word, lyrics) %>% 
  anti_join(stop_words) %>% 
  count(decade, word, sort = TRUE)

dtm <- word_counts %>% 
  cast_dtm(decade, word, n)
```

We then applied Latent Dirichlet Allocation (LDA). Although we initially experimented with multiple values for k (the number of topics), we found that increasing the number of clusters beyond two led to overly similar and repetitive topics that lacked meaningful separation. After testing various configurations, we determined that two topics (k = 2) provided the most interpretable and distinct division of themes across the corpus:

```{r}
lda_model <- LDA(dtm, k = 2, control = list(seed = 1234))
```

Next, we extracted the beta values from the model, which represent the probability of each word being associated with each topic. This allowed us to identify the top terms most representative of each topic:

```{r}
# Beta table: Probability of a word per topic
lda_topics <- tidy(lda_model)

# Top words per topic
top_terms <- lda_topics %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  ungroup() %>%
  arrange(topic, -beta)

top_terms
```

We visualized these top terms to gain an intuitive understanding of each topic’s content:

```{r}
# Plot
top_terms %>%
  mutate(term = reorder_within(term, beta, topic)) %>%
  ggplot(aes(beta, term, fill = as.factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  scale_y_reordered() +
  labs(
    title = "Top Words per Topic (LDA)",
    x = "Beta (Word Importance in Topic)",
    y = NULL
  )
```

Topic 1 is dominated by words such as **time**, **yeah**, **live**, **baby**, **heart**, and **feel**, suggesting a focus on personal emotion, relationships, and introspection. This cluster appears to capture the intimate and expressive side of songwriting, maybe reflecting themes of love, inner life, and emotional experience. The presence of colloquial markers like **yeah** and **gonna** further hints at a conversational and informal tone in the lyrics grouped under this theme.

Topic 2, on the other hand, centers around words like **night**, **life**, **break**, **play**, **money**, and **hold**. The inclusion of terms like night and play suggests a connection to nightlife or entertainment, while money and break might indicate ambition, success, or personal struggle. Compared to Topic 1, this cluster seems more outward-facing and grounded in external events or settings.

Overall, the LDA results offer a high-level view of the thematic duality in the lyrics: one topic capturing the emotional and relational aspects of songs, and the other reflecting experiential, material, and performative dimensions.

However, analyzing topic content alone wasn’t enough — we also wanted to understand how these topics varied in prominence over time. To do that, we extracted the gamma values from the model, which indicate how much each topic contributes to each document (in our case, decade or song). We linked each document back to its release decade to track changes in topic prevalence:

```{r}
# 1. Extract probabilities (gamma) for each document (song)
gamma_df <- tidy(lda_model, matrix = "gamma")  # lda_model = dein LDA-Modell

# 2. add release_date (so we can analyze decades)
# Assuming your tidy DTM is called `lyrics_dtm`
# and you have a column `release_date` in the original dataframe df

doc_info <- df %>%
  mutate(document = row_number(),              #  must match the sequence in the DTM!
         decade = floor(release_date / 10) * 10) %>%
  select(document, decade)

# 3. Link gamma values with decades
gamma_decade <- gamma_df %>%
  mutate(document = as.integer(document)) %>%
  left_join(doc_info, by = "document")
```

We then calculated the average topic proportion for each decade:

```{r}
# 4. Calculate average distribution of topics per decade
topic_by_decade <- gamma_decade %>%
  group_by(decade, topic) %>%
  summarize(mean_gamma = mean(gamma), .groups = "drop")
```

And plotted the temporal evolution of each topic’s prominence:

```{r}
# 5. Plot: Topic distribution per decade
ggplot(topic_by_decade, aes(x = decade, y = mean_gamma, fill = factor(topic))) +
  geom_col(position = "dodge") +
  labs(title = "Topic Prevalence per Decade",
       x = "Decade",
       y = "Average Gamma (Topic Proportion)",
       fill = "Topic") +
  theme_minimal()

```

```{r}
df <- df %>%
  mutate(song_id = row_number(),
         decade = floor(as.numeric(release_date) / 10) * 10)

# tokenization + stopwords
lyrics_words <- df %>%
  unnest_tokens(word, lyrics) %>%
  anti_join(stop_words)

# DTM with one document per song
lyrics_dtm <- lyrics_words %>%
  count(song_id, word) %>%
  cast_dtm(song_id, word, n)
```

```{r}
lyrics_lda <- LDA(lyrics_dtm, k = 2, control = list(seed = 1234))
```

```{r}
gamma_df <- tidy(lyrics_lda, matrix = "gamma") %>%
  mutate(song_id = as.integer(document)) %>%
  left_join(df %>% select(song_id, decade), by = "song_id")

```

```{r}
gamma_df %>%
  group_by(decade, topic) %>%
  summarise(avg_gamma = mean(gamma)) %>%
  ggplot(aes(x = decade, y = avg_gamma, fill = as.factor(topic))) +
  geom_col(position = "dodge") +
  labs(title = "Topic Prevalence per Decade",
       x = "Decade", y = "Average Gamma (Topic Proportion)", fill = "Topic")
```

# Conclusion

This project set out to explore how English-language song lyrics have evolved from the 1950s to the 2020s through a combination of text mining techniques. By analyzing emotional tone, linguistic structure, and thematic patterns, we observed a clear and consistent transformation in the content and mood of popular music.

Sentiment analysis revealed a steady shift from the optimism of early decades to a growing presence of negative and emotionally intense language. While joy remained a lyrical constant, emotions such as sadness, fear, and anger became increasingly dominant, particularly in recent decades. This suggests a move toward introspection and emotional complexity in songwriting.

The evolution of language—especially the rise of more aggressive or explicit terms—mirrored broader cultural and stylistic changes, such as the emergence of hip-hop and alternative genres. Bigrams and word networks further highlighted this shift, contrasting the romantic simplicity of earlier lyrics with the commercialism, social critique, and raw expression of modern music.

Topic modelling added a deeper layer to this narrative,( uncovering broader thematic shifts that point to a more diverse and nuanced musical landscape. Today’s lyrics reflect not only personal emotions but also collective concerns, identity, and performance (añadir lo que sea que añadamos con el topipc modelling)

In sum, the data show that song lyrics have not simply changed—they have matured. They are longer, more emotionally charged, more thematically varied, and more engaged with the world around them. Music, as a cultural artifact, continues to echo the times—and those echoes have grown loude and more complex.
